\vspace{-2cm}Temporal Congestion Analysis and Predictive Modeling of Ride-Hailing Trips in NYC Using High-Volume FHV Data

### 2025-04-21 paper.tex

\section{Abstract}
This paper explores traffic congestion patterns in New York City by analyzing High-Volume For-Hire Vehicle (FHV) trip records from services such as Uber and Lyft. Using data from all 12 months of 2024, we investigate how trip durations and speeds vary by time of day, day of week, and provider. After extensive pre-processing in Apache Spark (PySpark), we identify temporal trends and build predictive models in R using linear and logistic regression. The study demonstrates that the dataset size and analysis complexity necessitate big data tools for processing and modeling at scale.

\section{Introduction}
Urban traffic congestion continues to be a pervasive problem affecting the efficiency of transportation systems, contributing to environmental concerns, and impacting commuter well-being. In densely populated cities like New York, where ride-hailing services such as Uber and Lyft operate at massive scale, understanding when and where congestion occurs is critical for optimizing urban mobility.
This project investigates temporal patterns in congestion using High-Volume For-Hire Vehicle (FHV) trip data collected and published by the NYC Taxi and Limousine Commission (TLC). With over 200 million trip records from 2024 alone, the dataset provides a rich source of behavioral and operational insights. The central research question is: How do trip duration and speed, as proxies for congestion, vary across different times of day, days of the week, and ride-hailing providers?
The objectives of the study are to (1) identify temporal congestion trends through descriptive analysis, and (2) develop regression models to predict trip duration or classify trips as congested based on their features. Given the volume and granularity of the dataset, traditional analysis tools fall short, necessitating the use of big data processing technologies like Apache Spark. This research aims to produce interpretable insights that can inform urban planning, traffic policy, and ride-hailing operations.

\section{Background Review}

A growing body of research has explored the use of ride-hailing data for understanding urban transportation dynamics. Previous studies have focused on topics such as demand prediction, ride-sharing optimization, pricing strategy, and transportation equity. NYC’s Taxi and Limousine Commission has provided open access to large-scale trip data, which has enabled a range of analyses from temporal ridership patterns to spatial clustering of pick-up and drop-off zones \cite{nyctlc2023}.
Several works have leveraged this dataset to evaluate ride-hailing’s impact on traffic flow and public transit usage. For example, a study by NYU's Tandon School of Engineering found that New York City’s congestion surcharge reduced the number of ride-hailing trips but did not significantly alleviate congestion \cite{nyu2023}. Additionally, research from the CUNY Graduate Center examined the 2019 congestion surcharge’s effects on FHV usage in Manhattan, showing a decline in rides from charged zones \cite{cuny2023}.
Other studies have investigated surge pricing’s influence on driver behavior \cite{surgepricing2022} and analyzed pay disparities across platforms using data from 2019 to 2022 \cite{ucla2023}. However, many of these analyses are limited in scope—often examining only a few months of data, focusing exclusively on yellow taxi or green cab datasets, or lacking predictive components.
There remains a gap in literature concerning sustained, full-year temporal analysis of congestion based specifically on High-Volume FHV data. Furthermore, predictive modeling using such data is less common, particularly models focused on estimating congestion or trip delays. Our project addresses these shortcomings by utilizing a complete 12-month dataset and incorporating regression modeling. Additionally, our use of big data processing tools distinguishes this study from prior work that relies on single-machine or static data approaches.

\section{Method}
This study follows a multi-phase approach consistent with the big data lifecycle, including data acquisition, pre-processing, analysis, and modeling.

\subsection{Data Acquisition and Setup}
We obtained the NYC TLC High-Volume FHV trip record dataset for all twelve months of 2024. Each monthly dataset is stored in Parquet format and accessed via Google Drive in a cloud-based Google Colab environment. We used Apache Spark (PySpark) for all data ingestion and transformation steps, given the dataset's scale of 239,470,448 records.

\subsection{Data Processing}
In PySpark, we cleaned the data by removing rows with null or invalid values (e.g., zero or negative durations and distances). We then created derived columns for \texttt{trip\_duration\_min}, \texttt{avg\_speed\_mph}, and categorical features like \texttt{pickup\_hour}, \texttt{pickup\_day}, and \texttt{trip\_provider}. Outliers were filtered using interquartile range (IQR) thresholds for trip duration, distance, and speed. These pre-processing steps reduced noise and ensured reliable downstream analysis. Lastly, columns not needed for the analyses were removed from the dataset. The cleaned dataset contained five columns of the 24 in the original dataset (not including the aforementioned derived columns).

>>>>

\section{Abstract}
This paper explores traffic congestion patterns in New York City by analyzing High-Volume For-Hire Vehicle (FHV) trip records from services like Uber and Lyft. Using data from all 12 months of 2024, I investigate how trip durations and speeds vary by time of day, day of week, and provider. After extensive preprocessing in Apache Spark, I identify temporal trends and build predictive models in R using linear and logistic regression. The study demonstrates that the dataset size and analysis complexity necessitate big data tools for processing and modeling at scale.

\section{Introduction}
Urban traffic congestion continues to be a pervasive problem affecting the efficiency of transportation systems, contributing to environmental concerns, and impacting commuter well-being. In densely populated cities like New York, where ride-hailing services such as Uber and Lyftoperate at massive scale, understanding when and where congestion occurs is critical for optimizing urban mobility.

In this project, I investigate temporal patterns in congestion using High-Volume For-Hire Vehicle (FHV) trip data collected and published by the NYC Taxi and Limousine Commission (TLC). With over 200 million trip records from 2024 alone, the dataset provides a rich source of behavioral and operational insights. The central research question is: How do trip duration and speed, as proxies for congestion, vary across different times of day, days of the week, and ride-hailing providers?

The objectives of this study are to (1) identify temporal congestion trends through descriptive analysis, and (2) develop regression models to predict trip duration or classify trips as congested based on their features. Given the volume and granularity of the dataset, traditional analysis tools fall short, necessitating the use of big data processing technologies like Apache Spark. My goal is to produce interpretable insights that can inform urban planning, traffic policy, and ride-hailing operations.

\section{Background Review}
Recent work by Riley et al. proposes an integrated optimization and machine learning framework to improve real-time ride-sharing dispatch using NYC trip data, reducing average passenger wait times by over 50\% \cite{riley2020dispatching}. Similarly, Ma et al. introduce a congestion-aware dispatch and charging scheduling system for electric ride-hailing fleets, achieving improved profitability and service efficiency in high-congestion scenarios \cite{ma2024congestionaware}. These studies highlight advanced, data-driven approaches to mitigating ride-hailing congestion and support the potential for predictive modeling grounded in large-scale NYC mobility data.

A growing body of research has explored the use of ride-hailing data for understanding urban transportation dynamics. Previous studies have focused on topics such as demand prediction, ride-sharing optimization, pricing strategy, and transportation equity. NYC’s Taxi and Limousine Commission has provided open access to large-scale trip data, which has enabled a range of analyses from temporal ridership patterns to spatial clustering of pick-up and drop-off zones \cite{nyctlc2023}.

Several works have leveraged this dataset to evaluate ride-hailing’s impact on traffic flow and public transit usage. For example, a study by NYU Tandon School of Engineering found that New York City’s congestion surcharge reduced the number of ride-hailing trips but did not significantly alleviate congestion \cite{nyu2023}. Additionally, research from the CUNY Graduate Center examined the 2019 congestion surcharge’s effects on FHV usage in Manhattan, showing a decline in rides from charged zones \cite{cuny2023}.

Other studies have investigated surge pricing’s influence on driver behavior \cite{surgepricing2022} and analyzed pay disparities across platforms using data from 2019 to 2022 \cite{ucla2023}. However, many of these analyses are limited in scope—often examining only a few months of data, focusing exclusively on yellow taxi or green cab datasets, or lacking predictive components.

There remains a gap in literature concerning sustained, full-year temporal analysis of congestion based specifically on High-Volume FHV data. Furthermore, predictive modeling using such data is less common, particularly models focused on estimating congestion or trip delays. My project addresses these shortcomings by utilizing a complete 12-month dataset and incorporating regression modeling. Additionally, my use of big data processing tools distinguishes this study from prior work that relies on single-machine or static data approaches.

\section{Method}
This study follows a multi-phase approach consistent with the big data lifecycle, including data acquisition, preprocessing, analysis, and modeling.

\subsection{Data Acquisition and Setup}
I obtained the NYC TLC High-Volume FHV trip record dataset for all twelve months of 2024. Each monthly dataset is stored in Parquet format and accessed via Google Drive in a cloud-based Google Colab environment. I used Apache Spark (PySpark) for all data ingestion and transformation steps, given the dataset's scale of over 200 million records.

\subsection{Data Processing}
In PySpark, I cleaned the data by removing rows with null or invalid values (e.g., zero or negative durations and distances). I then created derived columns for trip_duration_min, avg_speed_mph, and categorical features like pickup_hour, pickup_day, and trip_provider. Outliers were filtered using interquartile range (IQR) thresholds for trip duration, distance, and speed. These preprocessing steps reduced noise and ensured reliable downstream analysis. After preprocessing, the dataset contained approximately 204 million valid trip records. Average trip duration was 18.2 minutes with a standard deviation of 12.3 minutes, while average trip distance was 4.8 miles. Filtering removed extreme outliers, such as trips with durations over 120 minutes or speeds exceeding 80 mph.
